---
title: "Exploration - CO2"
author: "Matthew Coad"
date: "2024-11-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(rprojroot)

source(rprojroot::is_rstudio_project$find_file("Scripts/Settings.R"))
source(climate_script_path("CO2.R"))

```

## IAC

[Historical CO2 Datasets](https://www.co2.earth/historical-co2-datasets). Like the temperature data its divided into historical reconstructions and more recent datasets based on direct instrument readings. 

Institute for Atmospheric and Climate Science (IAC) at Eidgenössische Technische Hochschule in Zürich, Switzerland.

Accessed from [Historical CO2 Datasets](https://www.co2.earth/historical-co2-datasets).
." for the historical data. I'll call this data set IAC.

### IAC Annual CO2 levels from 0 to 2014

```{r}
co2_iac_annual <- get_co2_iac_annual_0To2014()
glimpse(co2_iac_annual)
```
```{r, echo=FALSE, message=FALSE}
co2_iac_annual %>%
    ggplot(aes(x= Year, y = CO2.Global, color = CO2.Source)) + 
    geom_line() + 
    xlab("Year") + 
    ylab("Co2 Concentration (Co2)") +
    guides(color=guide_legend(title="Data Source")) +
    theme_bw()

```

### MAUNA

C. D. Keeling, S. C. Piper, R. B. Bacastow, M. Wahlen, T. P. Whorf, M. Heimann, and H. A. Meijer, Exchanges of atmospheric CO2 and 13CO2 with the terrestrial biosphere and oceans from 1978 to 2000. I. Global aspects, SIO Reference Series, No. 01-06, Scripps Institution of Oceanography, San Diego, 88 pages, 2001.

Dataset accessed 2018-09-03 at https://data.giss.nasa.gov/gistemp/.

### Mauna annual CO2 levels from 1959 present

```{r}
co2_mauna_annual <- get_co2_mauna_annual_1959toPresent()
co2_mauna_annual
```

```{r}
co2_mauna_annual %>%
    ggplot(aes(x= Year, y = CO2.Global, color = CO2.Source)) + 
    geom_line() + 
    xlab("Year") + 
    ylab("Co2 Concentration (Co2)") +
    theme_bw()

```

### CO2 Annual

Combined CO2 Annual data

```{r}
co2_annual <- get_co2_annual()
co2_annual
```
```{r}
co2_annual %>%
    ggplot(aes(x= Year, y = CO2.Global, color = CO2.Source)) + 
    geom_line() + 
    xlab("Year") + 
    ylab("Co2 Concentration (Co2)") + 
    labs(
        title = "Co2 Concentration from 2,000 years ago to present",
        caption = "Source: 'IAC' and 'MAUNA'") + 
    guides(color=guide_legend(title="Data Source")) +
    theme_bw()

```
