---
title: "R Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
library(knitr)
library(caret)
library(recipes)

knitr::opts_chunk$set(echo = FALSE)

source("./Scripts/Settings.R")
source("./Scripts/AusWeather.R")
source("./Scripts/AusClimate.R")
source("./Scripts/AusMaps.r")
source("./Scripts/AusRainfall.r")
source("./Scripts/SOI.r")
source("./Scripts/Utility.R")

```

## Load our best quality data

```{r}

# Analysis period
baseline_start <- 1950
baseline_end <- 2017
n_quantile <- 11

# Minimum number of years of data
# Allow stations to drop out for a few years
clean_years <- baseline_end - baseline_start - 2

# Load the data
rainfall_data <- aus_memorised_rainfall_data(baseline_start, baseline_end, clean_years)
stations <- rainfall_data$stations
observations <- rainfall_data$observations
monthly_observations <- aus_rainfall_monthly_observations(observations)
baselines <- aus_rainfall_baselines(monthly_observations, baseline_start, baseline_end, n_quantile)
monthly_valuations <- aus_rainfall_monthly_valuations(monthly_observations, baselines, n_quantile)
regions <- aus_rainfall_regions(stations, monthly_valuations, 2000, 10, 12, 3, 7)

soi_df <- soi_load() %>% mutate(SOI.Standard = scale(SOI))

```

## Build Model Input Data

```{r}

regional_df <- 
    monthly_valuations %>%
    inner_join(regions, by = c("Zone", "Station")) %>%
    group_by(Date, RegionID) %>%
    summarise(Rainfall = median(Rainfall), Rainfall.Rating = mean(Rainfall.Rating)) %>%
    group_by(RegionID) %>%
    mutate(Rainfall.Standard = scale(Rainfall), Rainfall.Rating.Standard = scale(Rainfall.Rating)) %>%
    inner_join(soi_df, by = "Date") %>%
    mutate(Year = year(Date), Month = month(Date)) %>%
    select(RegionID, Date, Year, Month, Rainfall, Rainfall.Standard, Rainfall.Rating, Rainfall.Rating.Standard, SOI, SOI.Standard)
regional_df %>% filter(RegionID == 3)

```

```{r}

regional_df %>%
    filter(RegionID == 3, Year >= 2010) %>%
    ggplot(aes(x = Date)) +
    geom_line(aes(y = Rainfall.Rating.Standard, color = "Rainfall")) +
    geom_line(aes(y = SOI.Standard, color = "SOI"), alpha = 0.6) +
    labs(
        x = "Date",
        y = "Values"
    ) +
    theme_bw()


```

## Run Training Experiment

```{r, fig.asp = 1}

# algorithms <- c("earth", "null", "svmRadial")
algorithms <- c("earth", "null")
lags <- seq2(0, 24)
tr_parameters <- 
    expand.grid(Algorithm = algorithms, N_Lag = lags) %>% 
    mutate(label = paste0("Algorithm ", Algorithm, " lag ", N_Lag))

regionID <- 3
tr_df <- function(n_lag) {
    regional_df %>% 
    filter(RegionID == regionID, Year >= baseline_start, Year <= baseline_end) %>%
    arrange(Date) %>%
    mutate(SOI.Lag = lag(SOI, n_lag), SOI.Standard.Lag = lag(SOI.Standard, n_lag)) %>%
    filter(!is.na(SOI.Lag))
}
    
tr_recipe <- function(n_lag) {
    recipe(Rainfall.Rating ~ SOI.Lag, tr_df(n_lag))
}

tr_control <- function() trainControl(method = "boot")
tr <- function(algorithm, n_lag) {
    set.seed(seed)
    train(tr_recipe(n_lag), tr_df(n_lag), method = algorithm, trControl = tr_control()) 
}
metric <- "RMSE"
seed <- 100
tr_emc <-
    emc_bind(
        tr_parameters,
        emc_record(
            train = tr(Algorithm, N_Lag),
            .label = label,
            .verbose = FALSE
        ),
        emc_performance(train)
    )
tr_working_emc <- tr_emc %>% filter(train_errors == 0)
tr_emc %>% select(Algorithm, N_Lag, train_error, train_warnings)

```

```{r}

tr_resamples <- 
    tr_working_emc %>% 
    emc_bind(emc_resamples(train)) %>% 
    unnest(resamples) %>% 
    select(Algorithm, N_Lag, RMSE, MAE, Resample)
ggplot(tr_resamples) + geom_boxplot(aes(x = Algorithm, y = RMSE)) + facet_wrap(~ N_Lag)

tr_performance  <- 
    tr_working_emc %>% 
    emc_bind(emc_performance(train)) %>% 
    select(Algorithm, N_Lag, RMSE, MAE)
ggplot(tr_performance) + geom_line(aes(x = N_Lag, y = RMSE, color=Algorithm))


```

```{r}

tr_df(0) %>%
    filter(Year >= 2010, Year <= 2020) %>%
    ggplot(aes(x = Date)) +
    geom_line(aes(y = Rainfall.Rating.Standard, color = "Rainfall")) +
    geom_line(aes(y = SOI.Standard.Lag, color = "SOI"), alpha = 0.6) +
    labs(
        x = "Date",
        y = "Values"
    ) +
    theme_bw()


```

```

```{r}

emc_augment <- function(.train, .data, .variable) {
    
    variable_name <- quo_name(.variable)
    prediction_var <- sym(sprintf("%s.Prediction", variable_name))
    residual_var <- sym(sprintf("%s.Residual", variable_name))
    train <- .train[[1]]
    if (is.null(train)) {
        return(tibble::tibble()[1,])
    }
    prediction <- predict(train, newdata = .data)
    original <- .data %>% pull(!! .variable)
    residual <- original - prediction
    augment_df <- bind_cols(!! prediction_var := prediction, !! residual_var := residual, .data)
    tibble(augment = list(augment_df))
}

tr_augment_emc <- 
    emc_bind(
        tr_working_emc,
        emc_augment(train, regional_df %>% filter(RegionID == regionID), quo(Rainfall))
    )
tr_augmented <- tr_augment_emc %>% select(algorithm, augment) %>% unnest(augment) %>% as_tibble()
tr_augmented

```

```{r}

algorithm <- "svmRadial"

regional_adjusted_df <- 
    tr_augmented %>% 
    filter(algorithm == "svmRadial") %>% 
    select(-algorithm) %>% 
    mutate(Rainfall.Adjusted = Rainfall.Residual + median(Rainfall)) %>%
    mutate(Rainfall.Adjusted.Standard = Rainfall.Residual / sd(Rainfall)) %>%
    mutate(RegionID = regionID) %>%
    select(
        RegionID, Date, Year, Month, 
        Rainfall, Rainfall.Standard, Rainfall.Rating, Rainfall.Adjusted, Rainfall.Adjusted.Standard, 
        SOI, SOI.Standard)
regional_adjusted_df


```

```{r}

regional_adjusted_df %>%
    filter(Year >= 2000) %>%
    ggplot(aes(x = Date)) +
    geom_line(aes(y = Rainfall, color = "Rainfall")) +
    geom_line(aes(y = Rainfall.Adjusted, color = "Adjusted Rainfall"), alpha = 0.6) +
    labs(
        x = "Date",
        y = "Rainfall"
    ) +
    theme_bw()


```
