---
title: "Rainfall by quartile exploration"
output: html_notebook
---

Exploration into analysing rainfall by dividing counts into quartiles.

Rainfall is highly variable, even in one location. Trying to analyse it on the basis of some sort of continuous range doesn't give a clear picture of whats going on.

Here we explore dividing rainfall into bins and seeing what we can see.

# Initialization

```{r message=FALSE}

library(knitr)

source("./Scripts/Settings.R")
source("./Scripts/AusWeather.R")
source("./Scripts/AusClimate.R")
source("./Scripts/AusMaps.R")
source("./Scripts/Utility.R")

```

## Load Data

```{r}

# Evaluation Settings

# Base start major period
baseline_start <- 1950
# Baseline end major period
baseline_end <- 2018
# Number of required clean years
clean_years <- 65
# Major period quosure
major_period <- aus_year_period()
# Minor period quosure
minor_period <- aus_monthly_period()
# Number of quartiles to use
n_quantile <- 11
quantile_offset <- ceiling(n_quantile/ 2)
# Center evaluation sequecne
evaluate_centers <- 4:10

valuation_info <- aus_rainfall_valuation_info
valuation_info$baseline_start <- baseline_start
valuation_info$baseline_end <- baseline_end
valuation_info$clean_years <- clean_years

baseline_stations <- aus_baseline_stations(valuation_info)

load_data <- function() {
    message("Loading Observations")
    observations <- aus_clean_observations(baseline_stations$Station, aus_rainfall_clean_info)
    list(
        observations = observations
    )
}

memorised_data <- aus_memorise(load_data, aus_rainfall_observations_info, paste0("RainfallBins v", aus_rainfall_version, " base from ", baseline_start, " to ", baseline_end, " min ", clean_years))

stations <- baseline_stations
observations <- memorised_data$observations
observation_info <- aus_rainfall_observations_info
observations
```

## Observations per period

```{r}

incomplete_period_observations <-
    observations %>% 
    mutate(Major_Period = !! major_period, Minor_Period = !! minor_period) %>%
    group_by(Zone, Station, Major_Period, Minor_Period) %>%
    summarise(
        Rainfall = sum(Rainfall, na.rm = TRUE),
        Rainfall.Source = first(Rainfall.Source)
    ) %>%
    ungroup() %>%
    arrange(Zone, Station, Major_Period, Minor_Period)

period_summary <-
    incomplete_period_observations %>% 
    group_by(Zone, Station, Minor_Period) %>%
    summarise(
        Rainfall.Mean = mean(Rainfall, na.rm = TRUE),
        Rainfall.Median = median(Rainfall, na.rm = TRUE),
        Rainfall.Min = min(Rainfall, na.rm = TRUE),
        Rainfall.Max = max(Rainfall, na.rm = TRUE),
        Rainfall.IQR = IQR(Rainfall, na.rm = TRUE)
    )

period_observations <-
    incomplete_period_observations %>% 
    group_by(Zone, Station) %>%
    complete(Major_Period = full_seq(Major_Period, 1), Minor_Period = 1:12) %>%
    ungroup() %>%
    inner_join(period_summary, by = c("Zone", "Station", "Minor_Period")) %>%
    mutate(
        Rainfall.Original = Rainfall,
        Rainfall = if_else(!is.na(Rainfall.Original), Rainfall.Original, Rainfall.Median)) %>%
    select(Zone, Station, Major_Period, Minor_Period, Rainfall, Rainfall.Original, Rainfall.Source) %>%
    arrange(Zone, Station, Major_Period, Minor_Period) %>%
    mutate(
        Rainfall.L3 = Rainfall + lag(Rainfall, 1) + lag(Rainfall, 2),
        Rainfall.L6 = Rainfall.L3 + lag(Rainfall, 3) + lag(Rainfall, 4) + lag(Rainfall, 5),
        Rainfall.L12 = Rainfall.L6 + lag(Rainfall, 6) + lag(Rainfall, 7) + lag(Rainfall, 8) + 
            lag(Rainfall, 9) + lag(Rainfall, 10)  + lag(Rainfall, 11),
        Rainfall.L24 = 
            Rainfall.L12 + 
            lag(Rainfall, 12) + lag(Rainfall, 13) + lag(Rainfall, 14) + lag(Rainfall, 15) + lag(Rainfall, 16) + lag(Rainfall, 17) + 
            lag(Rainfall, 18) + lag(Rainfall, 19) + lag(Rainfall, 20) + lag(Rainfall, 21) + lag(Rainfall, 22) + lag(Rainfall, 23) 
    )
period_observations
```

## Baselines

```{r}

# Make an open ended quantile by setting the first break to zero and the last break to infinity
open_quantile <- function(x) {
    x[1] <- 0
    x[length(x)] <- Inf
    x
}

# For all period observations in the baseline period work out summar stats for the current 3, 12 and 24 months
baselines <-
    period_observations %>% 
    filter(Major_Period >= baseline_start, Major_Period <= baseline_end) %>%
    group_by(Zone, Station, Minor_Period) %>%
    summarise(
        Rainfall.Max = max(Rainfall),
        Rainfall.Min = min(Rainfall),
        Rainfall.1st = quantile(Rainfall, 0.25),
        Rainfall.Median = median(Rainfall),
        Rainfall.3rd = quantile(Rainfall, 0.75),
        Rainfall.Quantiles = list(quantile(Rainfall, (0:n_quantile) / n_quantile, na.rm = TRUE)),
        Rainfall.OpenQuantiles = list(open_quantile(quantile(Rainfall, (0:n_quantile) / n_quantile, na.rm = TRUE))),
        
        Rainfall.L3.Max = max(Rainfall.L3),
        Rainfall.L3.Min = min(Rainfall.L3),
        Rainfall.L3.1st = quantile(Rainfall.L3, 0.25),
        Rainfall.L3.Median = median(Rainfall.L3),
        Rainfall.L3.3rd = quantile(Rainfall.L3, 0.75),
        Rainfall.L3.Quantiles = list(quantile(Rainfall.L3, (0:n_quantile) / n_quantile, na.rm = TRUE)),
        Rainfall.L3.OpenQuantiles = list(open_quantile(quantile(Rainfall.L3, (0:n_quantile) / n_quantile, na.rm = TRUE))),
        
        Rainfall.L12.Max = max(Rainfall.L12),
        Rainfall.L12.Min = min(Rainfall.L12),
        Rainfall.L12.1st = quantile(Rainfall.L12, 0.25),
        Rainfall.L12.Median = median(Rainfall.L12),
        Rainfall.L12.3rd = quantile(Rainfall.L12, 0.75),
        Rainfall.L12.Quantiles = list(quantile(Rainfall.L12, (0:n_quantile) / n_quantile, na.rm = TRUE)),
        Rainfall.L12.OpenQuantiles = list(open_quantile(quantile(Rainfall.L12, (0:n_quantile) / n_quantile, na.rm = TRUE))),
        
        Rainfall.L24.Max = max(Rainfall.L24),
        Rainfall.L24.Min = min(Rainfall.L24),
        Rainfall.L24.1st = quantile(Rainfall.L24, 0.25),
        Rainfall.L24.Median = median(Rainfall.L24),
        Rainfall.L24.3rd = quantile(Rainfall.L24, 0.75),
        Rainfall.L24.Quantiles = list(quantile(Rainfall.L24, (0:n_quantile) / n_quantile, na.rm = TRUE)),
        Rainfall.L24.OpenQuantiles = list(open_quantile(quantile(Rainfall.L24, (0:n_quantile) / n_quantile, na.rm = TRUE)))

    ) 
baselines
```

## Valuations

```{r}

valuations <-
    period_observations %>% 
    group_by(Zone, Station, Minor_Period) %>%
    inner_join(baselines, by=c("Zone", "Station", "Minor_Period")) %>%
    mutate(Date = make_date(Major_Period, Minor_Period)) %>%
    mutate(Rainfall.Rating = findInterval(Rainfall, first(Rainfall.OpenQuantiles)) - quantile_offset) %>%
    mutate(Rainfall.L3.Rating = findInterval(Rainfall.L3, first(Rainfall.L3.OpenQuantiles)) - quantile_offset) %>%
    mutate(Rainfall.L12.Rating = findInterval(Rainfall.L12, first(Rainfall.L12.OpenQuantiles)) - quantile_offset) %>%
    mutate(Rainfall.L24.Rating = findInterval(Rainfall.L24, first(Rainfall.L24.OpenQuantiles)) - quantile_offset) %>%
    ungroup() %>%
    select(
        Zone, Station, Major_Period, Minor_Period, Date,
        Rainfall.L12.Median,
        Rainfall, Rainfall.Rating,
        Rainfall.L3, Rainfall.L3.Rating,
        Rainfall.L12, Rainfall.L12.Rating,
        Rainfall.L24, Rainfall.L24.Rating, 
        Rainfall.Source)
valuations

```

## Rating plots

Plot a map of australia wide rainfall ratings

```{r}
selected_major_period <- 1970
selected_variable <- quo(Rainfall.L12.Rating)

for (selected_minor_period in 1:12) {
    plot(
        valuations %>%
        filter(Major_Period == selected_major_period, Minor_Period == selected_minor_period) %>%
        inner_join(stations, by = c("Zone", "Station")) %>%
        ggplot() %>%
        aus_dark_map() + 
        geom_point(aes(x = Longitude, y = Latitude, color = !! selected_variable)) +
        scale_colour_distiller(type = "div", palette="RdBu", direction = 1)) +
        ggtitle(paste0("Minor_Period = ", selected_minor_period))
}

```


# Regions

## Generate Options

```{r}

# Region generation options
region_baseline_start <- 2000
region_baseline_years <- 10
region_baseline_minor_period <- 12
distance_scale <- 3
region_baseline_end <- region_baseline_start + region_baseline_years - 1

# Spread the rainfall quantile during the baseline period into a row per station
region_valuations <-
    valuations %>% 
    filter(Major_Period >= region_baseline_start, Major_Period <= region_baseline_end, Minor_Period == region_baseline_minor_period) %>%
    inner_join(stations, by=c("Zone", "Station")) %>%
    mutate(Latitude = scale(Latitude) * distance_scale, Longitude = scale(Longitude) * distance_scale) %>%
    mutate(Rating_Value = scale(Rainfall.L3)) %>%
    select(Zone, Station, Latitude, Longitude, Rainfall = Major_Period, Rating_Value) %>% 
    spread(key = Rainfall, value = Rating_Value, sep = ".") %>%
    ungroup()
    
# Function that evaluates regions for a given number of Kmeans centers
regions_for_centers <- function(centers) {
    region_kmeans <- region_valuations %>% select(-Zone, -Station) %>% kmeans(centers)
    r <- region_valuations
    r$Centers <- centers
    r$RegionID <- region_kmeans$cluster
    r <- r %>% select(Zone, Station, Centers, RegionID)
    r
}

regions_by_centers <- evaluate_centers %>% map_dfr(regions_for_centers)
regions_by_centers

```

## Explore Options

Plot map of regions for each center

```{r}

for (center in evaluate_centers) {
    plot(regions_by_centers %>%
        inner_join(stations, by = c("Zone", "Station")) %>%
        filter(Centers == center) %>%
        ggplot() %>%
        aus_dark_map() + 
        geom_point(aes(x = Longitude, y = Latitude, color = RegionID)) +
        scale_colour_distiller(type = "div", palette="RdBu") +
        guides(colour=FALSE) +
        ggtitle(paste0("Centers = ", center)))
}


```

6 picks out Vic, NSW, Queensland/outback and wester AUS.

## Complete

```{r}
region_centers <- 7
station_regions <- 
    regions_by_centers %>% 
    filter(Centers == region_centers) %>% 
    select(Zone, Station, RegionID)

# Determine median longitude for each region
regionIDs <- station_regions %>%
    inner_join(stations, by = c("Zone", "Station")) %>%
    group_by(RegionID) %>%
    summarise(Median.Longitude = median(Longitude)) %>%
    ungroup() %>%
    arrange(Median.Longitude) %>%
    pull(RegionID)

region_valuations <- 
    valuations %>% 
    inner_join(station_regions, by = c("Zone", "Station")) %>% 
    select(Zone, Station, RegionID, everything())
region_valuations

```

## Plot ratio of stations in Drought

```{r, fig.asp = 1}

# start_period <- baseline_start
start_period <- 1950
end_period <- 2020
minor_period <- 12

layout <- rbind(c(1,1),
             c(2,3))

for (regionID in regionIDs) {
    
    df <- 
        region_valuations %>%
        filter(RegionID == regionID) %>%
        filter(Major_Period >= start_period, Major_Period <= end_period)
    df_summary <- 
        df  %>%
        group_by(Date) %>%
        summarise(Rainfall.L12.Median = median(Rainfall.L12))

    gridExtra::grid.arrange(
        
        df_summary %>%
            ggplot(aes(x = Date, y = Rainfall.L12.Median)) +
            geom_line(color = "grey30") +
            geom_smooth(formula = y ~ x, method = "lm", color = "blue", se = FALSE) +
            labs(
                x = "Year",
                y = "Rainfall for last 12 months (mm)"
            ) +
            theme_bw(),
        
        stations %>%
            inner_join(regions, by = c("Zone", "Station")) %>%
            filter(RegionID == regionID) %>%
            ggplot() %>%
            aus_map() + 
            geom_point(aes(x = Longitude, y = Latitude), size = 1) +
            guides(colour=FALSE) +
            labs(
                x = "Longitude",
                y = "Latitude"
            ) +
            theme_bw(),
        
        df %>%
            filter(Minor_Period == minor_period) %>%
            ggplot(aes(x = Date, y = Rainfall.L12)) +
            geom_point(aes(color = Station), alpha = .25, size = .75) +
            guides(colour=FALSE) + 
            coord_cartesian(ylim = c(0, 2000)) +
            labs(
                x = "Year",
                y = "Rainfall (mm)"
            ) +
            theme_bw(),

        layout_matrix = layout,
        heights= c(2,1)
    )
}


```



