---
title: "R Notebook"
output: html_notebook
---


Load scripts

```{r message=FALSE}

library(knitr)

source("./Scripts/Settings.R")
source("./Scripts/AusWeather.R")
source("./Scripts/AusClimate.R")
source("./Scripts/Utility.R")

library(recipes)

aus_map <- function(data) {
    
    map <- with_nameSpace(
        "maps", 
        ggplot2::borders("world", regions = "Australia",  fill = "grey90", colour = "black")
    )
    
    ggplot(data) + map  + xlim(112, 155) + ylim(-44, -10)  # Set y axis limits
}

aus_dark_map <- function(data) {
    
    map <- with_nameSpace(
        "maps", 
        ggplot2::borders("world", regions = "Australia",  fill = "DarkGrey", colour = "black")
    )
    
    ggplot(data) + map  + xlim(112, 155) + ylim(-44, -10)  # Set y axis limits
}

```

## Degenerate Stations

Get rainfall summary

```{r}

rainfall_complete_stations <- aus_load_stations(aus_rainfall_station_info)

rainfall_complete_summary <- 
    aus_observation_summaries(aus_rainfall_observations_info)

rainfall_complete_all_summary <- 
    rainfall_complete_summary %>% 
    unnest(All) %>%
    mutate('N %'= (N / (N_Year * 365)) * 100, `NA %` = (N_NA / N) * 100, 'Zero %' = (N_Zero / (N_Zero + N_NonZero)) * 100) %>%
    inner_join(select(rainfall_complete_stations, Zone, Station, Latitude, Longitude) , by = c("Zone","Station"))

rainfall_good_stations_code <-
    rainfall_complete_all_summary %>%
    filter(Status == "Good") %>%  # We actually need data
    filter(between(Longitude, 112, 155), between(Latitude, -44, -10)) %>% # Strip out all the remote islands.
    filter(N_NonZero > 0) %>% # If their are *no* recordings assume an outlier
    filter(N_Year >= 25) %>% # We need at least 20 years of data
    filter(`NA %` >= 25) %>% # We need more than 25% of data points filled
    pull(Station)

rainfall_good_stations <- rainfall_complete_stations %>% filter(Station %in% rainfall_good_stations_code)
rainfall_good_summary <- 
    rainfall_complete_summary %>% 
    filter(Station %in% rainfall_good_stations_code) %>%
    inner_join(rainfall_good_stations, by = c("Zone","Station")) %>%
    select(Zone, Station, Name, Latitude, Longitude, everything(), -variable, -units, -Status, -Error)

rainfall_good_all_summary <- 
    rainfall_good_summary %>% 
    unnest(All) %>% 
    select(-Decadal, -Yearly) %>% 
    mutate('N %'= (N / (N_Year * 365)) * 100, `NA %` = (N_NA / N) * 100, 'Zero %' = (N_Zero / (N_Zero + N_NonZero)) * 100)
rainfall_good_all_summary

```

List weather stations that are seem degenerate

```{r}

rainfall_complete_all_summary %>% 
    filter(NearZeroVar) %>% 
    select(Station, N_Year, N, `NA %`, 'Zero %') %>% 
    sample_n(20)

```

Some stations with a very low % of NAs have lots of zeros. Huge amounts of australia is arid. Maybe thats to be expected.

Plot Zero % on map.

```{r}

rainfall_complete_all_summary %>% 
    aus_map() + 
    geom_point(aes(x = Longitude, y = Latitude, color = `Zero %`), size = .5) +
    scale_colour_distiller(type = "div", palette="RdBu", direction = -1)
```

Stations recording high number of zeros are in arid regions. Recording large numbers of zeros is to be expected and is informational.

Plot Na % on map.

```{r}

rainfall_complete_all_summary %>% 
    aus_map() + 
    geom_point(aes(x = Longitude, y = Latitude, color = `NA %`), size = .5) +
    scale_colour_distiller(type = "div", palette="RdBu", direction = 1)
```

If we include all stations then we get that consistently across the board the majority of stations have a low number of measurements.
We'll drop out stations that record a low % of measurements.

```{r}

rainfall_good_all_summary %>% 
    aus_map() + 
    geom_point(aes(x = Longitude, y = Latitude, color = `NA %`), size = .5) +
    scale_colour_distiller(type = "div", palette="RdBu", direction = 1)
```

If we drop out stations with a low % of NAs then we get what seems to a scattering of stations across australia with a reasonble number in remote regions.

```{r}
rainfall_good_all_summary %>% 
    aus_dark_map() + 
    geom_point(aes(x = Longitude, y = Latitude, color = log2(Mean))) +
    scale_colour_distiller(type = "div", palette="RdBu", direction = 1)
```

## Samples Stations

Lets determine a random selection of stations to explore.

Pick a random set then use max dissimlarity to detect unusual samples.

Get a selection of 100 stations.


```{r}

# Choose a selection of statistics used to select a sample
sample_prep <- 
    rainfall_good_all_summary %>% 
    select(N_Year, Min_Year, Latitude, Longitude, N, 'N %','NA %', 'Zero %', Min, Mean, Max, SD, Skew)

# Center and scale all properties
sample_prep <- recipe(sample_prep) %>% step_center(everything()) %>% step_scale(everything()) %>% prep(retain = TRUE) %>% juice()

# Choose an initial random set of 5 and a further 15 records that have max disimilarity
set.seed(101)
sample_initial_idx <- sample(1:nrow(sample_prep), 5)
sample_dissim_idx <- caret::maxDissim(sample_prep[sample_initial_idx,], sample_prep, n = 20, verbose = FALSE)

sample_all_summary <- rainfall_good_all_summary[c(sample_initial_idx, sample_dissim_idx),]
sample_stations <- rainfall_complete_stations %>% filter(Station %in% sample_all_summary$Station)
sample_summary <- rainfall_good_summary %>% filter(Station %in% sample_all_summary$Station)
sample_all_summary <- sample_all_summary %>% select(Station, Latitude, Longitude, N_Year, Min_Year, Max_Year, N, 'N %', 'NA %', 'Zero %', Min, Mean, Max, SD, Skew)
sample_all_summary


```

Stations have measurments for almost every day they are considered to have a record for. Neat.

There are stations with no non-zero measurements. Dismiss them.
There are stations with only a few years of observations. Dismiss all stations with less than 20 years data.
There are stations that are almost entirely NA. Dismiss any with less than 25% of records.

# Sample Exploration

Plot locations of samples

```{r}

sample_all_summary %>% 
    aus_map() + 
    geom_point(aes(x = Longitude, y = Latitude, color = Mean)) +
    scale_colour_distiller(type = "div", palette="RdBu", direction = 1)


```


## Yearly Exploration

Plot number of observations per year across time.

```{r}

sample_complete_yearly_summary <- 
    sample_summary %>% 
    unnest(Yearly) %>% 
    mutate('N %'= (N / 365) * 100, `Present %` = (N_Present / N) * 100, 'Zero %' = (N_Zero / (N_Zero + N_NonZero)) * 100)

```

```{r}
sample_complete_yearly_summary %>% ggplot() + geom_col(aes(x = Year, y = N)) + facet_wrap(~ Station)
```

All the records are in solid blocks. There are no drop outs for stations in the records we have.

Plot % of NA of observations per year across time.

```{r}
sample_complete_yearly_summary %>% ggplot() + geom_col(aes(x = Year, y = `Present %`)) + facet_wrap(~ Station)
```

However the record keeping reliability of some stations is highly erratic. They often go through periods of not keeping data.

Plot % of Zero observations per year across time.

```{r}
sample_complete_yearly_summary %>% ggplot() + geom_col(aes(x = Year, y = `Zero %`)) + facet_wrap(~ Station)
```

Same story.The number of zeros recorded is inconsistent. Some years seem to have no data at all and thus we can't even calc a zero%.

Plot mean of observations per year across time. This is the sort of figure we will be interested in.

```{r}
sample_complete_yearly_summary %>% ggplot() + geom_col(aes(x = Year, y = Mean)) + facet_wrap(~ Station)
```

It looks way too erractic for some stations.

## Years with complete records.

To be included a year must have at least 25% of records recorded.

```{r}

sample_yearly_summary <- 
    sample_summary %>% 
    unnest(Yearly) %>% 
    mutate('N %'= (N / 365) * 100, `Present %` = (N_Present / N) * 100, 'Zero %' = (N_Zero / (N_Zero + N_NonZero)) * 100) %>%
    filter(`Present %` >= 25, N >= 360)

```

Now plot number of observations per year. We now expect gaps.

```{r}
sample_yearly_summary %>% ggplot() + geom_col(aes(x = Year, y = N)) + facet_wrap(~ Station)
```


Now plot % of Present observations per year across time. Years should an observable signal or be dropped out.

```{r}
sample_yearly_summary %>% ggplot() + geom_col(aes(x = Year, y = `Present %`)) + facet_wrap(~ Station)
```

Plot % of Zero observations per year across time.

```{r}
sample_yearly_summary %>% ggplot() + geom_col(aes(x = Year, y = `Zero %`)) + facet_wrap(~ Station)
```

Okay we have either a measure for a year or the year is dropped out entirely and the number of days per year where a zero was recorded looks consistent for each station over time.

A possibly exception if 96005.

Also some station have had their years almost entirely dropped out.

Really the important measure is mean rainfall per year.

```{r}
sample_yearly_summary %>% ggplot() + geom_col(aes(x = Year, y = Mean)) + facet_wrap(~ Station)
```

Lets drop station 31141 from our plot so we can focus on the remaining stations.

```{r}
sample_yearly_summary %>% filter(Station != '31141') %>% ggplot() + geom_col(aes(x = Year, y = Mean)) + facet_wrap(~ Station)
```

The means look more consistent over time but there may be some possible outliers.

Bouts of extreme weather are to be expected. Also we selected these stations as being odd.

## Stations with consistent records.

Lets filter our list down to stations that have a least 30 years of data where each year has at least 25% of days having an observation.

```{r}

consistent_year_cutoff <- 0.25

rainfall_station_consistency  <- 
    rainfall_good_summary %>%
    unnest(Yearly) %>% 
    mutate(Present_Ratio = N_Present / N) %>%
    filter(N > 350, Present_Ratio > consistent_year_cutoff) %>% 
    group_by(Zone, Station) %>% 
    summarise(N_consistent = n())

rainfall_consistency_summary <-
    rainfall_complete_summary %>%
    left_join(rainfall_station_consistency, by=c("Zone", "Station")) %>%
    unnest(All) %>% 
    mutate(N_Consistent = ifelse(!is.na(N_consistent), N_consistent, 0)) %>%
    mutate(Present_Ratio = N_Present / N) %>%
    select(
        Zone, Station, Status, N_Year, N_Consistent, Min_Year, Max_Year, 
        N, N_NA, N_Present, Present_Ratio)

rainfall_consistent_station_codes <- rainfall_consistency_summary %>% filter(N_Consistent > 30) %>% pull(Station)

rainfall_consistent_stations <- rainfall_complete_stations %>% filter(Station %in% rainfall_consistent_station_codes)

rainfall_consistent_all_summary <- 
    rainfall_complete_summary %>% 
    inner_join(rainfall_consistent_stations, by = c("Zone","Station")) %>%
    left_join(rainfall_station_consistency, by=c("Zone", "Station")) %>%
    mutate(N_Consistent = ifelse(!is.na(N_consistent), N_consistent, 0)) %>%
    unnest(All) %>% 
    select(-Decadal, -Yearly) %>% 
    mutate(Present_Ratio = N_Present / N) %>%
    select(Zone, Station, Name, Latitude, Longitude, 
           N_Year, N_Consistent, Min_Year, Max_Year,
           N, N_NA, N_Present, Present_Ratio, Min, Mean, Max, FirstQ, Median, ThirdQ, SD, Skew)

rainfall_consistent_all_summary
```

Yearly consistent summary

```{r}

rainfall_consistent_yearly_summary <- 
    rainfall_complete_summary %>% 
    inner_join(rainfall_consistent_stations, by = c("Zone","Station")) %>%
    left_join(rainfall_station_consistency, by=c("Zone", "Station")) %>%
    mutate(N_Consistent = ifelse(!is.na(N_consistent), N_consistent, 0)) %>%
    unnest(Yearly) %>% 
    mutate(Present_Ratio = N_Present / N) %>%
    filter(N > 350, Present_Ratio > consistent_year_cutoff) %>% 
    select(Zone, Station, Name, Latitude, Longitude, 
           Year, N_Year, N_Consistent, Min_Year, Max_Year,
           N, N_NA, N_Present, Present_Ratio, Min, Mean, Max, FirstQ, Median, ThirdQ, SD, Skew)

rainfall_consistent_yearly_summary
```

## Consistent exploration

Map of Log2 means

```{r}
rainfall_consistent_all_summary %>% 
    aus_dark_map() + 
    geom_point(aes(x = Longitude, y = Latitude, color = log2(Mean))) +
    scale_colour_distiller(type = "div", palette="RdBu", direction = 1)
```

I can see the rainfall by regions. You can see east of the great divde. Northern Australia, the deserts. Looking good.

## Exploration over time

Lets get a sample of consistent stations to explore.

```{r}
# Choose a selection of statistics used to select a sample
consistent_sample_prep <- 
    rainfall_consistent_all_summary %>% 
    select(N_Year, N_Consistent, Min_Year, N, Present_Ratio, Min, Mean, Max, SD, Skew)

# Center and scale all properties
consistent_sample_prep <- recipe(consistent_sample_prep) %>% step_center(everything()) %>% step_scale(everything()) %>% prep(retain = TRUE) %>% juice()

# Choose an initial random set of 5 and a further 15 records that have max disimilarity
set.seed(103)
consistent_sample_initial_idx <- sample(1:nrow(consistent_sample_prep), 5)
consistent_sample_dissim_idx <- caret::maxDissim(consistent_sample_prep[consistent_sample_initial_idx,], consistent_sample_prep, n = 20, verbose = FALSE)

consistent_sample_idx <- c(consistent_sample_initial_idx, consistent_sample_dissim_idx)

consistent_sample_all_summary <- rainfall_consistent_all_summary[consistent_sample_idx,]
consisten_sample_stations <- rainfall_complete_stations %>% filter(Station %in% consistent_sample_all_summary$Station)
consistent_sample_yearly_summary <- rainfall_consistent_yearly_summary %>% filter(Station %in% consistent_sample_all_summary$Station)

```

### Number of Records per year

```{r}
consistent_sample_yearly_summary %>% ggplot() + geom_col(aes(x = Year, y = N)) + facet_wrap(~ Station)
```

We now tend to have more complete records.

### Number of Observations per year

```{r}
consistent_sample_yearly_summary %>% ggplot() + geom_col(aes(x = Year, y = Present_Ratio)) + facet_wrap(~ Station)
```

Much more consistent presence ratio.

### Mean per year

```{r}
consistent_sample_yearly_summary %>% filter(Station != "31141") %>% ggplot() + geom_col(aes(x = Year, y = Mean)) + facet_wrap(~ Station)
```

